%%
%%  $Header: /u/gcmpack/MITgcm/pkg/frazil/frazil_description.tex,v 1.6 2012/03/11 19:41:27 dimitri Exp $
%%  $Name:  $
%%

Package ``FRAZIL'' is a repackaging and a replacement for subroutine
freeze_interior.F and for run-time option allowInteriorFreezing, which has
been retired.  It is meant to represent the formation of frazil at depth due
to pressure and freshening-related supercooling.

The initial functionality of pkg/frazil is similar to that previously provided
by freeze_interior.F, that is, it checks for water below the freezing point at
depth and brings the negative heat anomaly to the surface level in one time
step.  Specifically, pkg/frazil currently assumes that: (i) supercooling at
depth creates frazil, which rises to the model's surface level in one model
time step, (ii) the frazil column operates at sub-grid scale and rises to the
surface through levels warmer than freezing point without melting, and (iii)
the frazil is porous to salinity as it rises through the water column, that
is, frazil does not generate or transport salt anomalies.  All these
assumptions can be relaxed at a later time as the package matures.

The difference with freeze_interior.F is that rather than directly change
potential temperature THETA, pkg/frazil instead stores the required
adjustments in a frazil temperature forcing matrix:
FrazilForcingT (> 0 increases theta [W/m^2]).

The frazil forcing terms are computed prior to any of the ``ice'' packages and
routines (that is, seaice, thsice, shelfice, icefront, and freeze_surface).
This allows the ice packages to operate on FrazilForcingT directly (rather
than react to SST below the freezing point), if they need to do so, that is
they can operate on FrazilForcingT and decrease (or increase) it prior to
applying this FrazilForcingT to THETA in external_forcing.F

Package frazil is enabled with CPP option ALLOW_FRAZIL and runtime parameter
useFRAZIL, in the usual MITgcm ways.  There are currently no pkg/frazil
runtime options, so there is no need for data.frazil runtime option file.
